---
title: "Intro. to LA Using R"
subtitle: "Part 2: Foundational Skills"
author: "LASER Institute"
date: "Summer, 2021"
output:
  xaringan::moon_reader:
    css: [theme.css]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Welcome!

Welcome to *part 2*!

Great job so far.

---

# Foundational R skills

A general framework for you to use as a foundation and as a set of concepts to help you work through the institute.

The four core concepts we will use to build our framework are:

1. Projects
1. Packages
1. Data
1. Functions

You will use each of these in most of your analyses with R.

---

## 1. Projects

Working Directories 

```{r}
getwd() 
```

R Projects 

Project options are saved in .Rproj file

The here package helps you to navigate around in a project:

```{r, eval = FALSE}
library(here)

# data file
here("data", "fall=2018-data-file.csv")
```

---

# 2. Packages

1. What are packages? 
  + Code bundles that add functionality to R
  + Examples: ggplot2, dplyr, rtweet, quanteda, lme4
1. Where do we get packages?
  + CRAN or GitHub
1. How do we install packages?
  + install.packages("pkg-name")
1. How do we know what packages to use?
  + Searching 
  + People and news related to R (more later - there are *tons*)
  + CRAN task views
1. How do we use packages?
  + `library(pkgname)`

---

# 2. Installing another package

Tidyverse, a collection of R packages.

https://www.tidyverse.org/

Install via the following (do this now):

```{r, eval = FALSE}
install.packages("tidyverse")
```

What issues have arisen?

---

# 3. Data

So far, we have used *built-in data*. There is a lot of built-in data!

Loading different types of data

Comma-separated values (`.csv`)

```{r, eval = FALSE}
library(readr)
readr::read_csv(here("data", "filename.csv"))
```

---

# 3. Data 

`.xlsx`

```{r, eval = FALSE}
library(readxl)
read_excel((here("data", "schedule.xlsx")))
```

---

# 3. SPSS

`.sav`

```{r, eval = FALSE}
library(haven)
read_sav((here("data:, file-name.sav")))
```

---

# 3. Other data sources

Google Sheets

```{r, eval = FALSE}
library(googlesheets4)
```

Web

```{r, eval = FALSE}
read_csv("https://github.com/data-edu/dataedu/raw/master/data-raw/wt01_online-science-motivation/raw/s12-course-data.csv")
```

---

# 4. Functions

- A function is a reusable piece of code that allows us to consistently repeat a programming task

- Functions in R can be identified by a word followed by a set of parentheses, like so: `word()`. 

More often than not, the word is a verb, such as `filter()`, suggesting that we’re about to perform an action. 

Indeed, functions act like verbs: they tell R what to do with our data.

The parentheses are where we can provide arguments.

---

# 4. Functions

- What is the name of the *package* used below?
- What is the name of the *data* used below?
- What is the name of the function used below?*

```{r, eval = FALSE}
library(dplyr)
#mtcars
glimpse(mtcars)
```

---

# 4. `select()`

```{r, message = FALSE}
library(dplyr)

storms %>% 
  select(name, year, month, day, hour, status)
```

---

# 4. `filter()`

```{r}
library(dplyr)

storms %>% 
  filter(month == 8)
```

---

# 4. `arrange()`

```{r}
library(dplyr)

storms %>% 
  arrange(hour)
```

---

# 4. Putting it together

```{r}
library(dplyr)

storms %>% 
  select(name, year, month, day, hour, status) %>% 
  filter(month == 8) %>% 
  arrange(hour)
```

---

# 4. Assignment operator

```{r}
storms_in_august <- storms %>% 
  select(name, year, month, day, hour, status) %>% 
  filter(month == 8) %>% 
  arrange(hour)
```

What is one thing that is different between `storms_in_august` and `storms` ?

---

# 4. Assignment operator

```{r}
storms
```

---

# 4. Assignment operator

```{r}
storms_in_august
```

---

# 4. Assignment operator

```{r}
ncol(storms)
```

Tricky question: How many columns are present in `storms` after the following 
operation?

```{r, eval = FALSE}
storms %>% 
  select(name, year, status) 
```

---

# 4. Assignment operators

How many columns are in storms after running the following two lines of code?

```{r, eval = FALSE}
storms <- storms %>% 
  select(name, year, status) 
```

---

# 4. Pipe operator

We've been using the pipe operator `%>%` from the magrittr package

The pipe sends the results of a function (or object) from left side of pipe to next function after pipe.

So instead of this:

```{r, eval = FALSE}
library(magrittr)
library(dplyr)

mtfilter <- dplyr::filter(mtcars, mpg < 20)
mtsubset <- dplyr::select(mtfilter, mpg, cyl, disp)
```

We can do this:

```{r, eval = FALSE }
mtcars %>% 
  dplyr::filter(mtcars, mpg < 20) %>% 
  dplyr::select(mpg, cyl, disp)
```


---

# 4. Basic programming operators

Math

```{r}
x <- 3 * 4
x
y <- 2 + 3 -1
y
z <- 4 / 2 ** 3
z
```

---

# 4. Basic programming operators

Logic

```{r}
x <- TRUE
y <- FALSE
# and
x & y
# or 
x | y
# not 
!x
```

---

# Discuss in groups (or, if there is insufficient time, in Slack)

- What is one thing you learned from this part?
- What questions do you still have?
