---
title: "Learning Analytics Foundations Learning Lab 3"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
name: ''
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this learning lab, we'll build on the visualizations we created (and
the wrangling we had to do to be in the position to create our
visualizations!) by creating and sharing a report of our work. In that
report, we'll include interactive tables - tables that can help us and
others to *explore* our data.

## 1. PREPARE

### Packages 📦

We'll load the same packages we've loaded in the past two learning labs.

#### [**Your Turn**]{style="color: green;"} **⤵**

This time, though, it's your turn to the code to load the three packages
we have loaded and used: {tidyverse}, {janitor}, and {here} in the code
chunk below.

In addition please load three additional packages that we have not used
*yet*, but will use in this lab. These packages are {skimr}, {corrr},
and {apaTables}. All three are already installed. In sum, you'll be
loading six packages - more than we've used so far, but not atypical for
a complex analysis.

```{r}
library(tidyverse)
library(janitor)
library(here)
library(skimr)
library(corrr)
library(apaTables)
```

### Data

We'll load the `joined_data` data frame we created in the last learning
lab. If you saved it, great. We also prepared a copy and saved it in the
`/data` directory.

We can load, or read, that file with the following code:

```{r}
joined_data <- read_csv(here("data", "joined-data.csv"))
```

#### Data Source \#3: Self-Report Survey

The third data source is a self-report survey. This was data collected
before the start of the course. The survey included ten items, each
corresponding to one of three motivation measures: interest, utility
value, and perceived competence. These were chosen for their alignment
with one way to think about students' motivation, to what extent they
expect to do well (corresponding to their perceived competence) and
their value for what they are learning (corresponding to their interest
and utility value). We'll use this in the third learning lab.

We'll read this file, named `survey.csv`, which is---like our other data
files---stored in the `/data` directory.

#### [**Your Turn**]{style="color: green;"} **⤵**

In the code below, read the `survey.csv` file. You can use the code
above as a template. Assign the output from the `read_csv()` function
the name `survey`.

```{r}
survey <- read_csv(here("data", "survey.csv"))
```

**Hint**: By asking you to "assign the output from the `read_csv()`
function the name `survey`", consider how in the code chunk above this
"Your Turn" code chunk, we assigned the output from the `read_csv()`
function to the name `joined_data`.

After reading the data, let's continue the practice of looking at our
data. Type `survey` into the console to take a look at the data: Does it
appear to be the correct file? What do the variables seem to be about?
What wrangling steps do we need to take? Taking a quick peak at the data
helps us to begin to formulate answers to these and is an important step
in any data analysis, especially as we *prepare* for what we are going
to do.

Add one or more of the things you notice or wonder about the data here:

-   

We'll take one more step with this data frame. We could this in the code
above or after; they're different means of reaching the same end,
cleaning up these names a bit.

```{r}
survey <- clean_names(survey)
```

Let's take one more look at the data by typing its name into the console
to check that the above function appeared to work; if it did, the names
should be lower-case, and any symbols or spaces should now be replaced
by an underscore (`_`).

A small next step. Because we used `clean_names()` with the survey data,
we need to make the `Subject`, `Course`, and `Section` variable names
in`joined_data` lower-cased. We can quickly use `clean_names()` on this
data frame or can `rename()` to do this:

```{r}
joined_data <- joined_data %>% 
  rename(subject = Subject,
         semester = Semester,
         section = Section)
```

## 2. WRANGLE

You may have noticed there is something fishy in three columns in the
data that contain "date" in their names. The suffix `_x` and `_y` (or
`.x` and `.y` if we do not use `clean_names()`) typically suggest that
there was an issue at some point when data files were joined; if a
column is present in both of the files we are joining, then they will be
identified in the file output by the join by these suffixes.

In short, these variables are fishy. Moreover, since this this data
represents students' responses to a survey at the beginning of the
semester, it may be that the specific date and time at which they
completed the survey is unimportant.

Let's clean up a bit by removing these extraneous (for this but perhaps
not for *every* analysis) variables. To do so, we'll use a `select()`
"helper" function. You may recollect the `select()` function from the
Getting Started task; it's back! As you saw in that task, we can use
`select()` to select (in a particular order) variables in your data
frame. This is often helpful, as there are many techniques for which
having a focused data set can be quite helpful (we'll say more on these
later; one such case is when you're transforming your data from "long"
to "wide" format).

In addition to what `select()` does at its core, helper functions can
make select even more useful. Let's begin by continuing our practice of
looking at our data; either print `survey` to the console or run the
`glimpse()` function with the survey data (e.g., type and enter
`glimpse(survey)` in the console) to see what variables exist in the
file (you may need to expand your console a bit to view them all!).

One helper function is relatively simple; using the `:` symbol. This
lets us select all of the variables *in between* two variables. For
instance, we may wish to select all of the variables between
`student_ID` and `section`, which we can do as follows:

```{r}
survey %>% 
  select(student_id:section)
```

The above code outputs a data frame with far fewer columns - five
instead of 26! Just in case this is something you're thinking about now,
the data frame that is printed above is not yet *saved* in any way; it's
simply printed to the console. How might we save (or assign) the data
frame? If we may want to use any of the 21 variables we did not select
in another analysis, we might choose to assign the output a different
name than survey---that way, both `survey` and `survey_subset` are
present in our environment, and we can choose to use either based on our
specific purpose.

*Aside:* It is common to create subsets of data in this way; the new
data frame can be named whatever you like, including `survey_subset` as
well as `mickey_mouse`! Of course, like naming Google Docs files and
data files on our computer, we often wish to choose brief, informative
names, better yet within R if they're relatively easy to type, which
typically means names that have only lower case letters and have no
spaces or symbols apart from underscores (which happens to be the
convention used by `clean_names()`.

```{r}
survey_subset <- survey %>% 
  select(student_id:section)
```

But, back to selecting.

#### [**Your Turn**]{style="color: green;"} **⤵**

Below, make two modifications to the above code:

1.  Select not all of the variables between `student_id` and `section`,
    but instead all of the variables between `student_id` and `percomp`.
2.  Assign a name other than `survey_subset` to the output. To help to
    assign an informative name, it may be helpful to know that `int`
    refers to a composite measure of students' self-reported interest in
    science; `val` refers to their self-reported value for science; and.
    `percomp` refers to their degree of confidence in doing well in
    their science class.

```{r}
my_survey_questions <- select(student_id:percomp)
```

After you've run the code above, take a quick peak at the data one more
time by typing (or `glimpse()`'ing) the new data frame you've created.

Another useful select helper function is the minus symbol, `-`.

For instance, we can remove the `date` variable from our data set
through the use of the following code:

```{r}
survey %>% 
  select(-date)
```

The minus symbol (`-`) can also be combined with other functions.
Imagine that we wanted to remove all three of the date-related
variables---deciding that they're not relevant to our analysis. We can
do this with the following code:

```{r}
survey %>% 
  select(-contains("date"))
```

Contains selects (or, if it is prefaced by a minus symbol, removes) all
variables that contain the string you specify. This can be helpful when
specific variables in your data share a common name. For instance, you
may analyze data from survey software that denotes questions from a
similar group or section of your survey with a particular name or term.
`contains()` can help you to select all of the variables that contain
that name or term.

Let's proceed to selecting a subset of the survey data to use. We need
the `student_id` and the three variables related to the course
(`subject`, `semester`, and `section`), as well as the three variables
that we described above that relate to students' motivation: `int`,
`val`, and `percomp`. We'll select those by using the colon (`:`) symbol
as well as selecting the `student_id` variable.

```{r}
survey_to_join <- survey %>% 
  select(student_id, subject:percomp)
```

At this point, we likely have a number of different data frames in our
environment, and that's okay! We'll focus on two from here forward:
`joined_data` and `survey_to_join`. You are likely anticipating a
sensible next step: Let's join these together so we have one data frame
we can explore further!

This should be a straightforward join, just like we did in the last
learning laboratory. Let's go ahead and join the two files as follows:

```{r}
# this returns an error! Read on
full_join(joined_data, survey_to_join)
```

Oof! That was an unexpected error. Sometimes, it is difficult to figure
out what R is telling us when we see an error. Other times, we can begin
to scrutinize (or Google!) the error message to gain a sense for what
didn't work quite as we expected. Here, the error is:

> Error: Can't join on \`x\$student_id\` x \`y\$student_id\` because of
> incompatible types. ℹ \`x\$student_id\` is of type \<double\>\>. ℹ
> \`y\$student_id\` is of type \<character\>\>.

What do you think this error is telling us?? Debugging is as much an art
(and a creative process) as it is systematic. Often, we're looking for
hints or clues to tell us what to try out next. Do you have any hints
about either what is not working---or what we can do? Add one or more
below:

-   

You may be anticipating a way forward here. Reading this, we can see
that one of the variables that would serve as a *key* on which we would
join together our two data frames, `joined_data` and `survey_to_join` ,
has "incompatible types." Recall that the join functions we are using
like for any *key* variables to be strings. In other words, they cannot
be numbers, but must be words. Often, though, we wish for a key to be a
number---perhaps when an anonymous ID for a teacher or student is a
number. In these cases, we are not out of luck, but we need to tell R
that we want for what it thinks is a number to be a character string. In
other words, we want to change the type of a numeric variable to be a
character string variable.

Which of the two `student_id` variables is a number? One potential way
to determine this is to print both data frames to the console and to
check the variables. The error also presented us with a clue,
specifically this part:

> ℹ \`x\$student_id\` is of type \<double\>\>. ℹ \`y\$student_id\` is of
> type \<character\>\>.

These are output in the order we entered them, so the first of the two
lines above refers to the first data frame we used in the join function
and the second line refers to the second data frame. Since we ran
`full_join(joined_data, survey_to_join)`, `joined_data` is first and
`survey_to_join` is second. In other words, the number that needs to be
a character string is in the `joined_data` file. Let's change that to a
character string by using the `mutate()` function used yesterday.

```{r}
joined_data <- joined_data %>% 
  mutate(student_id = as.character(student_id))
```

Finally, let's try to join the data once more

```{r}
# after changing student_id to be a character string in both data frames
# the two data frames join without an error
full_join(joined_data, survey_to_join)
```

#### [**Your Turn**]{style="color: green;"} **⤵**

Above, we did not save the output from the join to a new object, in part
as we were checking to make sure the join worked as intended. Below,
join together `joined_data` and `survey_to_join` using `full_join()`,
assigning the name `data_to_explore` to the output.

```{r}
data_to_explore <- full_join(joined_data, survey_to_join)
```

## 3. EXPLORE

At this point, we should have quite the comprehensive data set,
including single measures from a) students for the time they spent in
the course LMS and other information about them, such as information on
why they are enrolled in the course, b) their academic achievement

We'll explore our data in two ways---by creating:

1.  Descriptive statistics for key variables in our data
2.  A correlation matrix for key variables that are numbers

We'll take these in turn, considering two different ways to create
correlation tables that may be suited better to particular tasks
depending on one's goals.

#### Creating descriptive statistics for key variables in our data

An efficient package for creating descriptive statistics *when your goal
is to understand your data internally* (rather than to create a table
for an external-to-the-research-team audience, like for a manuscript) is
the {skimr} package. A key feature of the {skimr} package is that it
works well with the {tidyverse} packages we are using: it takes data
frames as input, and returns data frames as output, which means we can
manipulate them with {tidyverse} functions (with `select()`, `filter()`,
and `arrange()`, for example).

It's best to dive right into our data with this, so let's go!

#### [**Your Turn**]{style="color: green;"} **⤵**

The challenge here is not the complexity of the `skim()` function, per
se, but will be comprehending the terminology. In the code chunk below:

-   Pass to the `skim()` function a single argument (recall from our
    brief day 1 presentation that functions have *names* and
    *arguments!*)

-   That single *argument* is the data frame (aka in tidyverse parlance,
    a tibble) for which you are aiming to calculate descriptive
    statistics

```{r}
skim(data_to_explore)
```

**Hint:** For help, consider running `?skim()` in the console to view
the documentation for the function.

**Hint**: If you see an error that says the function could not be found,
ensure that the {skimr} package is loaded (with `library(skimr)` earlier
in this document.

What do you notice about the output? These observations might pertain to
the format of the output or its values (i.e., what the mean for the
`val` variable is). Note one or two of these noticings or wonderings
below:

-   

As we noted earlier, the {skimr} package works nicely with other
{tidyverse} functions.

#### [**Your Turn**]{style="color: green;"} **⤵**

Recall from the Getting Started task how we *composed* a function,
step-by-step. Imagine we wish to remove the variables
`total_points_possible` and `total_points_earned` from our `skim()`
summary; can you do this by modifying the code below? Consider using the
`contains()` helper function for this use case!

```{r}
data_to_explore %>% 
  select(proportion_earned, time_spent, int, val, percomp) %>% 
  skim()
```

We noted earlier that this output is best for internal use. This is
because the output is rich, but not well-suited to exporting to a table
that you add, for instance, to a Google Docs or Microsoft Word
manuscript. Of course, these values can be entered manually into a
table, but we'll also discuss ways later on to create tables that are
ready--or nearly-ready--to be added directly to manuscripts.

If you are curious about doing more with {skimr}, we'll include a link
in the reach section at the end of this learning lab.

#### Creating a correlation matrix for key variables that are numbers

There are two efficient ways to create correlation matrices, one that is
best for internal use, and one that is best for inclusion in a
manuscript.

First, the {corrr} package provides a way to create a correlation matrix
in a {tidyverse}-friendly way. Like for the {skimr} package, it can take
as little as a line of code to create a correlation matrix. If not
familiar, a correlation matrix is a table that presents how *all of the
variables* are related to *all of the other variables*.

#### [**Your Turn**]{style="color: green;"} **⤵**

One key is to correlate only numeric variables. `select()` can help us
with this. Below, choose the variables you wish to correlate (consider
selecting at least 4-5), selecting only the numeric variables by adding
to the `select()` function. While some numeric variables can
*technically* be used, it is likely not sensible to correlate all of the
variables; some---for instance, the `section` variable---are not very
sensible to correlate!

```{r}
data_to_explore %>% 
  select(proportion_earned, time_spent, int, val, percomp) %>% 
  correlate()
```

#### [**Your Turn**]{style="color: green;"} **⤵**

Imagine we selected the following three variables. Replace those three
with the variables you selected, and examine how the results are
modified to a) include only the variables "below the diagonal" of the
correlation matrix---because, as you might have observed above, those
above and below the diagonal are mirror images of the other---and b) to
be rounded to two decimal points to be easier to read.

```{r}
data_to_explore %>% 
  select(int, val, percomp) %>% 
  correlate() %>% 
  shave() %>% 
  fashion()
```

While {corrr} is a nice package to quickly create a correlation matrix,
you may wish to create one that is ready to be added directly to a
manuscript. {apaTables} is great for creating more formal forms of
output that can be added directly to an APA-formatted manuscript; it
also has functionality for regression and other types of model output.
It is not as friendly to {tidyverse} functions; first, we need to select
only the variables we wish to correlate.

Then, we can use that subset of the variables as the argument to
the`apa.cor.table()` function.

#### [**Your Turn**]{style="color: green;"} **⤵**

Below, create a subset of the larger `data_to_explore` data frame with
the variables you wish to correlate, then create a correlation table
using `apa.cor.table()`.

```{r}
data_to_explore_subset <- data_to_explore %>% 
  select(int, val, percomp)

apa.cor.table(data_to_explore)
```

This may look nice, but how to actually add this into a manuscript? Here
is perhaps the most challenging task yet---perhaps well suited for the
final "Your Turn" of this learning lab. Read the documentation for
`apa.cor.table()` by running `?apa.cor.table()` in the console. Look
through the documentation and examples to understand how to output a
file with the formatted correlation table, and then run the code to do
that with your subset of the `data_to_explore` data frame.

```{r}
apa.cor.table(data_to_explore, filename = "survey-cor-table.doc")
```

**Hint**: Consider the `filename` argument to the `apa.cor.table()`
function.

## 5. COMMUNICATE

In this learning lab, we continued to go deeper into wrangling data,
doing so in advance of joining together our data frame we created in the
last learning lab with a third data frame including students'
self-reports of their interest, value, and confidence in science. Then,
we created descriptive statistics and correlation matrices, using
different packages for each.

To complete this learning lab, "knit" the document to HTML. You can view
the file in your web browser by clicking the name of the file that is
generated (it will end in `.html`) after the file runs (it may take a
few moments to render)

Good work! Below, add a few notes in response to the prompts:

*One thing I took away from this learning lab*:

*One thing I want to learn more about*:

## Reach

For this learning lab, create descriptive statistics for a data set of
your choosing *using one of the three packages we used in this lab*.

If you do not have data you wish to join at this time, consider reading
one of these resources on the three packages we used in this learning
lab:

-   Using skimr:
    <https://cran.r-project.org/web/packages/skimr/vignettes/skimr.html>

-   Correlations in R: <https://corrr.tidymodels.org/>

-   apaTables:
    <https://www.rdocumentation.org/packages/apaTables/versions/2.0.8>
