---
title: "Learning Analytics Foundations Learning Lab 1"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
name: ''
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, please add your name within the single quotation marks above -
this document is yours!

## Introduction

In this learning lab - one of sixteen throughout the two-weeks of the
summer LASER institute - we'll be working on two skills that in some
ways typify a learning analytics approach to working with data. These
two skills are:

1.  Wrangling data to be in the form we need to create a visualization
2.  Creating and beginning to style a complex plot - namely, a faceted
    plot

What makes these two things distinctive?

First, joining together different data sets is certainly a feature of
other research approaches. However, it can be difficult to do in some
software programs. Learning analytics research is in many ways typified
by the need to join together *many* different data sets, including
demographic or institutional record data, data from a course or learning
management system, and self-report questionnaire or survey data.

Second, faceted plots are on of the features of using R that can show
the benefits of using a programming language. Faceted plots can reveal a
great deal about your data in relatively short order. Learning to facet
plots can also teach a great deal about visualizing data using R.

Without further delay, let's go!

## 1. PREPARE

Like in the getting started task, we often prepare by establishing the
goals of our analysis and what techniques and data we'll need to achieve
that goal. Before we prepare in this way, though, we have a task that
will help us communicate throughout the institute. In particular, we'll
first get setup using Slack, a communication platform for teams and
groups like all of us who are involved with the LASER Institute.

### Slack

*Especially* if you haven't used Slack before, please \#tryitout! Even
if you have used Slack before, please complete these steps; if you have
used Slack before, they probably will not take you long to complete. 

-   First, follow this link to create an account and login to LASER
    Institute Scholars workspace (or to use an existing account to login
    to the workspace):
    <https://join.slack.com/t/laser-scholars/shared_invite/zt-rg2wstp0-rSlXEQoramDWYT1ZVpV~MA>

-   Consider downloading the desktop app now and logging in - it can be
    helpful, especially when it comes to quickly navigating the app and
    receiving notifications.

-   Next, find the \#tryitout channel (this should be visible in the
    left side of the LASER Institute Scholars workspace). Channels are
    used to organize communication about particular topics, like project
    meetings or presentations.

-   In the \#tryitout channel, *write a short introduction*. You can use
    this template. "Hi \#tryitout! I am [insert your first name as you
    wish to be called] and I am a [insert position at the University].
    One thing you might not know about me is that [insert a fact about
    yourself here]". 

-   Here's an example: "Hi \#tryitout! I am Josh and I am from the
    University of Tennessee, Knoxville. One thing you might not know
    about me is that I like to make pickles."

-   Next, find the post from someone else in \#tryitout, and add a
    *reaction* to their post. You can do this by hovering over their
    post (in the desktop app) or pressing down on someone else's post
    (in the mobile app) and then selecting a reaction, like the
    sunglasses emoji.

-   Finally, add a picture to your slack account (from computer): Please
    go to the left side of your slack account and click on your name.
    Then select 'profile and account'. Go to the 'edit profile' (on the
    right side of your workspace) and upload an image.

We will use Slack throughout the institute as a way to ask and answer
others' questions. We'll discuss how to most effectively solicit and
provide help in the presentation before the next learning lab.

If you've completed these steps, you're now done with \#tryitout in
Slack! Please do not hesitate to contact us for help at any step of the
way.

### Packages 📦

For this learning lab, we'll use the {tidyverse} set of packages--like
we used in the getting started task. We'll also use one another package,
{[janitor](https://sfirke.github.io/janitor/)}, which can help
particularly with the wrangling part of the LASER cycle. We'll use one
last package, {[here](https://here.r-lib.org/index.html)}, which helps
us to specify the paths to files in a way that will work well across
computers and even operating systems---your own and others'.

```{r}
library(tidyverse)
library(janitor)
library(here)
```

### Data

For this and the other three learning labs in this module, we'll be
using the online science classes data.

The online science classes we explore in this chapter were designed and
taught by instructors through a state-wide online course provider
designed to supplement---but not replace---students' enrollment in their
local school. For example, students may have chosen to enroll in an
online physics class because one was not offered at their school.

The data were originally collected for a research study, which utilized
a number of different data sources to understand students'
course-related motivation. These data sets included:

1.  *Log data*, data output from the learning management system (LMS) on
    the time students spent logged in to the LMS
2.  *Academic achievement* data on the number of points students earned
    (and how many points it was possible for them to achieve)
3.  *Self-report survey* data from a survey of students at the beginning
    of the semester.

In this Learning Analytics Foundation module, we'll be examining these
three data sets, and how key variables (such as the time students spent
on the LMS and the percentage of points they earned) might differ across
subjects (physics, biology, anatomy and physiology, oceanography, and
forensic science).

The first data set---log data---is what we'll describe and present here.
We'll describe the other data sets in the next three learning labs.
We'll also load the one we'll be using in this learning lab, the log
data. Recall from the Getting Started task that to run code in a code
chunk, we can click the green arrow to the far right. We'll also discuss
other means of running code in the presentation in advance of the next
learning lab.

#### Data Source \#1: Log Data

*Log-trace data* is data generated from our interactions with digital
technologies, such as archived data from social media postings. In
education, an increasingly common source of log-trace data is that
generated from interactions with LMS and other digital tools.

The data we will use is a summary type of log-trace data: the number of
minutes students spent on the course. While this data type is fairly
straightforward, there are even more complex sources of log-trace data
out there (e.g., time stamps associated with when students started and
stopped accessing the course). We'll name this data set `time_spent`, to
help us to quickly recollect what function it serves in this analysis.

```{r}
time_spent <- read_csv(here("data", "log-data.csv")) %>% 
  clean_names()
```

Type `time_spent` into the console (below this window) and then hit
return/enter. You should see a printed summary of this data frame. What
do you notice about this data? What do you wonder? Add a note (or
more---you can type return/enter after a bullet point to add another) on
your noticings and wonderings here:

-   

This is the only data set we will be using in this first learning lab in
this module.

## 2. WRANGLE

There are two steps that we will take to wrangle the data.

In particular, we'll be:

1.  Processing the `course_id` variable
2.  Mutating a column to change the time spent variable to represent
    hours

In doing these, we'll gain experience with data wrangling skills that
we'll expand upon in the other learning labs that make up the
Foundational Skills module---as well as in the other modules that make
up the institute.

### Processing the course_id variable

Earlier, we loaded `time_spent`, which contains information on the
number of minutes that students spent on the course, as well as other
variables, particularly `course_id`.

Information about the course subject, semester, and section are stored
in a single variable---`course_id`. This format of data storage is not
ideal. If we instead give each piece of information its own column,
we'll have more opportunities for later analysis. We'll use a function
called `separate()` to do this.

First, let's practice with a small data set. We'll create it directly in
R; run the code below to do that (and to assign the name `df` to the
dataset).

```{r}
df <- tibble(course_var = c("Fall - Chemistry", 
                            "Fall - Earth Science", 
                            "Spring - Forensic Science",
                            "Spring - Earth Science",
                            "Spring - Biology"))
```

Print `df` to the console. You should see a single variable,
`course_info`, with four rows.

In this (very small) data frame, there is information about both the
semester and the course are encoded within the same variable. The
`separate()` function has two primary arguments, one each for:

1.  the variable you want to separate
2.  the names of the new variables to create

Below, see using `course_var` for \#1, and `c("Semester", "Course")` for
\#2, can be used to separate the semester and course data into two
separate variable

```{r}
df %>% 
  separate(course_var, c("Semester", "Course"))
```

Next, let's try something slightly different. Here, we have a data frame
with a variable that encodes three pieces of information within the same
variable: the year, semester, and subject. There are also a few other
differences.

```{r}
df2 <- tibble(course_variable = c("19-Fall-Algebra I", 
                                  "20-Fall-Algebra II", 
                                  "20-Spring-Algebra I",
                                  "20-Spring-Algebra II",
                                  "21-Fall-Algebra I"))
```

#### [**Your Turn**]{style="color: green;"} **⤵**

Can you separate the variable in the above data frame not into two, but
rather three, new variables? Below is some template code.

```{r}
df2 %>% 
  separate(course_variable,
           c("Year", "Semester", "Sibject"))
```

**Hint**: Try to modify the code from above (in which you separated
`course_var` into two variables) based on a) the name of the variable in
`df2` and b) adding the name for the third new variable you wish to
create.

#### [**Your Turn**]{style="color: green;"} **⤵**

You're up again! Let's return back to our `time_spent` data frame. It is
often helpful to take a look at the data before writing code.

Below, we will load `time_spent` and run the `separate()` function with
the `course_id` variable to split up the subject, semester, and section
so we can use them later on. In other words, whereas above we separated
the variable `course_variable`, in the data set we'll use here, we'll
separate the `course_id` variable.

```{r}
time_spent %>%
  separate(course_id,
           c("Subject", "Semester", "Section"))
```

There is one last key step---one that is likely to be a bit disorienting
at first---that we'll do next. Once we've processed the data how we
would like, we have to assign, or save, the results back to the name for
the data with which we have been working. This is done with the
assignment operator, or the \<- symbol. Copy the code you successfully
ran in the chunk above to follow the assignment operator in the chunk
below. In other words, write the code you wrote above, but assign the
output back to `time_spent`.

```{r}
time_spent <- time_spent %>%
  separate(course_id,
           c("Subject", "Semester", "Section"))
```

We have made a habit of continually looking at our data after running
code to ensure that the step worked as intended. Not in a code chunk,
but rather in the console below, type the name of the data we have been
working with to ensure that the course_id variable has been separated
into three variables that correspond to the subject, semester, and
section.

If those look good, let's proceed to the next step. If something doesn't
look right, consider re-running the code chunks above, perhaps returning
all the way to the first code chunk that you ran (to load the data) to
ensure that the output is as you intended for it to be.

### Mutating a column to change the time spent variable to represent hours

In the above code, you used separate to create new variables based on an
existing variable. While that function serves a specific problem (when
there are effectively multiple variables combined in one), its use
represents a pattern that is fairly common: you use a function to solve
a problem; figuring out how it works, checking the output, then
assigning the output back to the name of the data frame, after which you
can proceed to the next step.

There are a lot of other functions like separate that help you to solve
specific problems, and we'll introduce many over the two-week institute
- and will point you to resources that describe many more.

There are also functions that can serve as general purpose tools that
can solve many problems; one of the most useful is `` mutate()` ``, a
function to create new variables in a data set. Specifically, we'll use
`mutate()` to create a new variable for the percentage of points each
student earned; keep in mind as you work through these steps how so many
parts of wrangling data involves either changing a variable or creating
a new one. For these purposes, mutate can be very helpful.

Let's begin again with a small data set with two variables, `var_a` and
`var_b`. Run the chunk below.

```{r}
df3 <- tibble(var_a = c(30, 50, 30, 10, 30, 40, 40, 30, 20, 50),
              var_b = c(100, 90, 60, 70, 60, 80, 70, 50, 30, 20))
```

Next, print `df3` to the console. You should see two numeric variables;
imagine they represent points that students earned on a 50-point quiz
and a 100-point test, respectively. There are a lot of things that you
might wish to do with these variables. For instance, you may wish to sum
them together. The code below does this.

```{r}
df3 %>% 
  mutate(points_sum = var_a + var_b)
```

#### [**Your Turn**]{style="color: green;"} **⤵**

We can combine many `mutate()` functions together. Below, create a new
variable (let's call it `points_proportion`) that represents the
proportion of the total points students could, potentially, earn. To do
this, you can divide `points_sum` by the maximum possible points---150.

```{r}
df3 %>% 
  mutate(points_sum = var_a + var_b) %>% 
  mutate(points_proportion = points_sum / 150)
```

**Hint**: Just like you can use the + symbol to add variables together,
you can use the / symbol to divide a variable by another---or by a
value, like 150!

After adding the above, you should see output that contains four
variables, one each for `var_a` and `var_b`, `points_sum`, which
represents the sum of the points students earned, and
`points_proportion`, which represents the percentage of the total points
students earned.

#### [**Your Turn**]{style="color: green;"} **⤵**

When we left off with `time_spent`, we had processed the `course_id`
variable. Let's now process the `time_spent` variable. Specifically,
this variable currently represents the number of *minutes* that students
spent on the course LMS. Below, add to the template of code below to
create a new variable, `time_spent_hours`, that represents the number of
*hours* that students spent on the course LMS.

```{r}
time_spent %>% 
  mutate(time_spent_hours = time_spent_minutes / 60)
```

**Hint**: Refer to the code you wrote above, being clear about a) what
the name of the new variable you are creating is and b) how you will
create this variable using division (by the number of minutes in an
hour).

We used the above as a test bed to ensure that our code worked as
intended. Once we are confident that we are creating the variable in the
way we intend to, we can assign the output back to the data frame that
`time_spent` refers to.

```{r}
time_spent <- time_spent %>% 
  mutate(time_spent_hours = time_spent_minutes / 60)
```

Good work wrangling this dataset!

## 3. EXPLORE

We're now ready to create a faceted plot. Like in the getting started
task, we'll use the ggplot2 package.

The code below creates a histogram with 30 bins---the default number for
`geom_histogram`. Change the number of bins below and note any
differences in what you interpret about the data.

```{r}
time_spent %>% 
  ggplot(aes(x = time_spent_hours)) +
  geom_histogram(bins = 30)
```

What do you think the ideal number of bins is---with what is ideal being
the number of bins that helps you to interpret the overall distribution
of the values for how much time students' spent (*note*: there is no one
right or wrong answer here!)?

-   

We'll next be using the `facet_wrap()` function to create *small
multiples*, or plots that are specific to subsets of your data. These
subsets are identified based on another variable in your dataset. For
example, the code below uses the built-in `mpg` dataset to plot the
relationship between the displacement of a car's engine and its highway
miles per gallon fuel efficiency.

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point()
```

The code in the next plot creates individual plots for each
class---think compact car or SUV.

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  facet_wrap(~class)
```

#### [**Your Turn**]{style="color: green;"} **⤵**

In the code below, create a faceted histogram based on the subject of
the course. To do so, consider both:

-   What code you used to create the histogram of the time students'
    spent on the course

-   How, in the example above, `facet_wrap` refers to the variable in
    that data frame that represents the class of the car---but modifying
    the code to work for your subject variable

You may also wish to change the color; reflect back to the getting
started task for an example of how to do this.

```{r}
ggplot(time_spent, aes(x = time_spent_hours)) + 
  geom_point() +
  facet_wrap(~Subject)
```

What do you notice about this figure? And what do you wonder? Add a note
(or a few notes!) below:

-   

## 5. COMMUNICATE

In this learning lab, we used focused on learning about and wrangling
data from an LMS on the time students spent on the course. Using
`separate()` and the generally helpful `mutate()` function, we processed
a data set such that we could then later create a plot that depicted the
distribution of a single variable---for how much time students' spent on
the course in hours by subject. Below, add a few notes in response to
the following prompts.

*One thing I took away from this learning lab:*

*One thing I want to learn more about:*

If you've made it to this point, congratulations - you've completed the
first learning lab! To complete your work, you can click the drop down
arrow at the top of the file, then select "Knit top HTML". This will
create a report in your Files pane that serves as a record of your code
and its output you can open or share.

If you wanted, you could save the processed data set to your data
folder. The `write_csv()` function is useful for this. The following
code is set to not run, as we wanted to ensure that everyone had the
data set needed to begin the second learning lab, but if you're
confident in your prepared data, you can save it with the following:

```{r, eval = FALSE}
write_csv()
```

## Reach (Optional)

Each learning lab will also contain a "reach," a chance for you to do
more if you wish to apply what you have learned to data for a project,
or to simply do more with a technique.

For this learning lab, your reach is to create a faceted plot (using
`facet_wrap()`) for your own data). You can do this below in this file
if you wish; you can add code chunks by clicking the green button that
includes a plus symbol and a letter "c" in the upper-right corner of
this window (to the left of the "Run" button).

If you do not have data with which to create a faceted plot, another
option is to click "Tutorial" in the top right corner of your RStudio
window and to begin one of the {learnr} tutorials. If you start one of
these, take a note on this work here as a record of what you've began.

In short, please (optionally) start on this work right here, including
reading data, preparing it, and then creating a faceted plot using data
of your choosing.

If you do this, re-knit your document when complete or at a stopping
point so you have a record of your work!
