---
title: "Learning Analytics Foundations Learning Lab 3"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
name: ''
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this learning lab, we'll build on the visualizations we created (and the wrangling we had
to do to be in the position to create our visualizations!) by creating and
sharing a report of our work. In that report, we'll include interactive tables - 
tables that can help us and others to _explore_ our data.

## 1. PREPARE

### Packages 📦

We'll load the same packages we've loaded in the past two learning labs. This time, though,
it's your turn to the code to load the three packages we have loaded and used: {tidyverse}, {janitor},
and {here} in the code chunk below.

```{r}

```

### Data

We'll also load one additional data source.

#### Data Source #3: Self-Report Survey

The third data source is a self-report survey. This was data collected before the start of the course. The survey included ten items, each corresponding to one of three motivation measures: interest, utility value, and perceived competence. These were chosen for their alignment with one way to think about students' motivation, to what extent they expect to do well (corresponding to their perceived competence) and their value for what they are learning (corresponding to their interest and utility value). We'll use this in the third learning lab.

## 2. WRANGLE

#### [**Your Turn**]{style="color: green;"} **⤵**

## 3. EXPLORE

#### [**Your Turn**]{style="color: green;"} **⤵**

## 4. MODEL

#### [**Your Turn**]{style="color: green;"} **⤵**

## 5. COMMUNICATE

#### [**Your Turn**]{style="color: green;"} **⤵**