---
title: "Learning Analytics Foundations Learning Lab 2"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
name: ''
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Building on the last learning lab in which we accessed data on the time
students spent on the LMS across courses in five different subjects, in
this learning lab, we'll add a second data set to the mix. This data set
is on the points students earned in the course.

In addition to joining together the two datasets (based on a common
*key*), we'll examine the relations between these two variables - time
spent on the course and points earned.

## 1. PREPARE

### Packages 📦

We will load the same three data sets we loaded in the first learning
lab: {tidyverse}, {janitor}, and {here}.

```{r}
library(tidyverse)
library(janitor)
library(here)
```

### Data

#### Data Source \#2: Academic Achievement Data

In addition to the `time_spent` data we loaded We'll explain a bit more
about the second data set - on academic achievement data.

Academic achievement data is (obviously) is a very common form of data
in education. In this learning lab, we'll use both the sum of the points
students earned as well as the number of points possible to compute the
percentage of points they earned in the course---a measure comparable
(but likely a little different based on teachers' grading policies) to
their final grade. We'll use this in the second learning lab.

We'll load the data in the same way as earlier:

```{r}
gradebook <- read_csv(here("data", "gradebook-summary.csv")) %>% 
  clean_names()
```

You may choose to type the name of the gradebook into the console window
and to then (like earlier) type enter/return to view a summary of what
data this name points to.

We'll also add the *wrangled* version of the data set we used in
learning lab \#1. This file should be saved in the data directory and is
named "log-wrangled.csv".

#### [**Your Turn**]{style="color: green;"} **⤵**

Your turn! Using the code in the code chunk above as a starting point,
modify `read_csv()` to read in the file "log-wrangled.csv".

```{r}
log_wrangled <- read_csv()
```

In the environment tab of your RStudio Cloud window, you should now see
the two data sets (gradebook and log_wrangled) loaded. If so, let's
proceed to the next steps. If you need any help, then reach out to us
via Slack.

## 2. WRANGLE

If our goal is to join together these data sets to explore how the time
spent on the course is related to the final grades achieved by students,
then we need to take a few steps, first, to ensure that the files are
ready to be joined. In particular, we'll be:

1.  Cleaning up the student ID variable in the log data

2.  Processing the gradebook data

3.  Joining the two data frame together into one data frame

Let's take these in turn.

### Cleaning up the ID variable in the log data

To join two datasets, it is important that the *key* (or *keys*) on
which you are joining the data is formatted identically. The key
represents an identifier that is present in both of the data sets you
are joining. For instance, you may have data collected from (or created
about) the same students that are from two very different sources, such
as a self-report survey of students and their teacher-assigned grades in
class.

In the present case, we have gradebook data and data from the LMS on the
time students spent on the course, but in one case---for the log
data---the student identifier is formatted a bit differently. This is
often the case, and to ensure that we can join together the two datasets
into one, we'll have to clean up the student_id variable, first. Let's
take a look at the log data by running the code in the chunk below.

```{r}
log_data
```

You may notice something is a bit different with this variable compared
to the gradebook data - the student_id is prefaced with an underscore
that the LMS happened to include.

We can remove these underscores using a technique in R for working with
text - techniques we'll explore in greater depth in a later module. For
now, we'll be using the str_sub function (that is a part of the
tidyverse). This function takes as input a string, or a word---or a
variable with a string---and returns a subset of the word based on the
position of the letters. There are lots of big ideas here, so let's walk
through an example.

Run the code below, in which we specify that we want the output to begin
with the first letter of the string, and to end with the tenth letter.
In other words, we want to return the entire word - something that is
not useful just yet.

```{r}
str_sub("Task 1 %54", start = 1, end = 10)
```

Now, let's imagine that our data erroneously included the characters
"%54", and that we wanted to remove those from our string so that what
was returned was only the task and the number.

#### [**Your Turn**]{style="color: green;"} **⤵**

Can you modify the code below so that the output returned is only the
task and the number?

**Hint**: Try changing the values entered for the start and end
arguments to the function.

```{r}
str_sub("Task 1 %54", start = 1, end = 10)
```

A feature of str_sub is that it not only works for a single string, but
also a collection of strings. Run the code below with the start and end
values you entered above to clean up a collection--- technically a
vector created with the combine, or c(), function---of strings next.

```{r}
str_sub(c("Task 1 %54",
          "Task 2 %54",
          "Task 3 %54",
          "Task 4 %54",
          "Task 5 %54",
          "Task 6 %54",
          "Task 7 %54",
          "Task 8 %54",
          "Task 9 %54",
          "Task 10 %54"),
        start = 1, end = 10)
```

Detour complete! Let's now consider how we can use what we did above to
clean the student_id variable in the log_data file.

In the code below, a template that returns the entire student ID is
provided.

#### [**Your Turn**]{style="color: green;"} **⤵**

How can you modify this code to remove the leading underscore? Note that
in addition to using mutate to create a new version of our student_id
variable, in the next line, we change the student_id variable to be an
integer. This is because our ID variable in the other data set is an
integer, and variables not only need to be named identically but also
need to be the same data type in order to be joined together.

**Hint**: Like above, consider changing the start and end values.

```{r}
log_wrangled %>% 
  mutate(student_id = str_sub(student_id, start = 1, end = 7)) %>% 
  mutate(student_id = as.integer(student_id))
```

Note that a bit different from in the example above, here we are using
str_sub() *within* a mutate function. In other words, whereas earlier we
cleaned a vector of strings (consisting of ten strings for Tasks 1-10),
in the code above, we're still using str_sub() with a vector, but it is
a vector representing *all* of the student_id variables in our log_data.

Whereas the above code printed the output for you to view, the code
below assigns---saves--- the output to log_data, such that when you use
that data file in subsequent steps, the changes you made will be
reflected. In other words, the student_id variable will be cleaned and
ready to go for the next steps.

#### [**Your Turn**]{style="color: green;"} **⤵**

Below, make the same change to the start and/or end values you made in
the previous chunk and then run the chunk.

```{r}
log_wrangled <- log_wrangled %>% 
  mutate(student_id = str_sub(student_id, start = 1, end = 7)) %>% 
  mutate(student_id = as.integer(student_id))
```

### Processing the gradebook data

In the previous step, we further processed the log data. In this step,
we'll process the gradebook data. In particular, we'll separate the
`course_id` variable in the same way we separated that variable in the
log data, and we'll also calculate a new variable representing the
proportion of points students earned (out of the points possible to
earn).

Let's start with separating the `course_id` variable. Run the code in
the next chunk to do this. If you named the three parts of the course ID
differently than they're named below (and saved the data you processed
to use in this learning lab), be sure that these three variables are
named identically; this is the *key* (pun intended!) to these variables
joining correctly.

```{r}
gradebook <- gradebook %>% 
  separate(course_id,
           c("Subject", "Course", "Section"))
```

Next, we'll mutate our data set to create a new column, one representing
the proportion of points students earned. Let's consider a data frame
with example data, `df4` .

```{r}
df4 <- tibble(var_a = c(8, 8, 7, 8, 9, 6, 8, 8, 7, 8),
              var_b = 9)
```

*Note*: To create `df4`, for `var_a`, we passed a *vector* that we
created with the function `c()` that contains 10 values. Consider these
to be the number of times that learners participated in an
outside-of-school STEM club. Instead of passing another vector for
`var_b`, we simply used the value 9, which represents the number of
opportunities students had to participate in the outside-of--school STEM
club . In this case, the value 9 is repeated for however many rows there
are in the data frame. Thus, in the context of creating a data frame,
`var_b = 9` is the same as `var_b = c(9, 9, 9, 9, 9, 9, 9, 9, 9, 9)`.

Since interpreting proportions when the denominator is nine can be
difficult, we may which to create a variable for the proportion.

#### [**Your Turn**]{style="color: green;"} **⤵**

After running the chunk above, print `df4` to the console to get a sense
of what the data frame consists of. To create a third variable that
represents the proportion of STEM club activities students participated
in, divide `var_a` by `var_b`.

```{r}
df4 <- df4 %>% 
  mutate()
```

**Hint**: Take a peak at your work on the last learning lab as an
example.

What happens if the output is different than you intended? That's no
problem! Re-run the code-chunk *above* (in which we create `df`) to have
a blank slate with which to try again.

#### [**Your Turn**]{style="color: green;"} **⤵**

Your turn once more. This time, create a new variable---here, let's name
it `proportion_earned`---using the `gradebook` data. This will involve
using the mutate function with the `gradebook` data, creating a new
variable (`proportion_earned`) on the basis of the values of two
existing variables:

-   `total_points_possible`

-   `total_points_earned`

Also, once your code is ready, you'll need to assign the results back to
`gradebook`. This is challenging as you're starting from scratch with
the code. However, good R programmers use other code (that they or
others wrote!) often, so feel free to copy and paste code from other,
similar problems to give yourself a head start.

```{r}

```

Once the above step is complete, take another look at `gradebook` by
printing it to the console. There should now be seven columns, the six
originally in the data and a new, seventh variable you've mutated.

### Joining the data

We're now ready to join! At their core, joins involve operations on two
data frames at the same time. This may seem useful only in certain
cases, but consider the following data analysis tasks:

-   You have collected data from *students* from one of ten classrooms;
    at the same time, you have data on the *teachers* of those ten
    classes (five of which tried out a new curriculum, and five who
    taught a "business-as-usual" curriculum

-   You are studying the posts on *Twitter* and *Pinterest* of one of
    around 100 mathematics teachers

-   After working with a local school district, you collected survey
    responses from 100s of *teachers* who teach in one of approximately
    25 elementary, middle, and high *schools*; you received data from
    the district on the characteristics of the schools, including how
    many students they serve and how many teachers work in them

In each of these cases---and many others like them---your single
analysis involves multiple data files. While in some cases it is
possible to analyze each data set individually, it is often useful (or
necessary, depending upon your goal) to join these sources of data
together. This is especially the case for learning analytics research,
in which researchers and analysts often are interested in understanding
teaching and learning through the lens of multiple data sources,
including digital data, institutional records, and survey data, among
other sources. In all of these cases, knowing how to promptly join
together files---even files with tens of thousands of hundreds of
thousands of rows---can be empowering.

Enough on the magic and beauty of joins; let's try a few out!

Consider two example data frames. `df5` contains a variable with four
student names, `name` and a variable for the number of STEM-related
classes they have taken, `n_stem_classes`.

`df6` contains a variable with three student names, `name` (like in
`df5`), as well as another, different variable, for students'
self-reported interest in STEM topics, `interest_in_stem`, measured on a
one-seven scale, with seven indicating higher levels of interest.

Run the code below and then type `df5` and `df6` in the console.

```{r}
df5 <- tibble(name = c("Sheila", "Tayla", "Marcus"),
              n_stem_classes = c(4, 5, 6))

df6 <- tibble(name = c("Tayla", "Marcus", "Sheila", "Vin"),
              interest_in_stem = c(4, 7, 6, 6))
```

A key (there's that pun, again) with joins is to consider what
variable(s) will serve as the *key*. This is the variable to join by. A
key must have two characteristics; it is:

-   a character string--- a word (thus, you cannot join on a number
    unless you "coerce" or change it to be a character string, first)

-   present in both of the data frames you are joining.

While some of the time it takes some thought to determine what the key
is (or what the keys are---you can join on multiple keys!), in the above
case, there is just one variable that meets both of the above
characteristics.

#### [**Your Turn**]{style="color: green;"} **⤵**

In the code below, enter the name of the variable that is the key within
the quotation marks following `by =`. Then run the code chunk and note
the output.

```{r}
full_join(df5, df6, by = "")
```

What do you notice about the output of the `full_join()`? All
observations are valid; consider how the output is similar to and
different from `df5` and `df6`, particularly in one or more notes
following the bullet point.

-   

`full_join()` is one of a number of joins from which we can choose.
`full_join()` is distinguished from the other joins by how it returns
*all* of the rows in both of the data frames being joined. If a
particular key is present in one of the data frames but not the other,
the values for the variable in the data set for which the key is not
present are simply recorded as missing (like in the above, where there
is no value for the number of STEM classes Vin has taken).

There is one other join on which we'll focus for now. That is
`left_join()`, which differs from `full_join()` in that it returns all
of the rows in the "left" data frame, the data frame named first in the
function, but not all of the rows in the "right" data frame: it retains
only the rows in the "right" data frame, the data frame named second in
the function, that have a matching key. An example is necessary. Before
running the code below, add the same key you added above.

#### [**Your Turn**]{style="color: green;"} **⤵**

```{r}
left_join(df5, df6, by = "")
```

Different from the above, `left_join()` did not return all of the rows
from both data frames, instead returning *all* of the rows in the "left"
data frame (and those in the "right" data frame with a match).

For now, we're going to use a single join function, `full_join()`. In
the code below, join `gradebook` and `log_wrangled`; type the names of
those two data frames as arguments to the `full_join()` function in a
similar manner as in the `full_join()` code above, and then run this
code chunk. For now, don't specify anything for the `by =` part of the
function.

```{r}
# join together the gradebook and log_wrangled
joined_data <- full_join()

joined_data
```

You may notice a red message that says
`Joining, by = c("student_id", "Course", "Subject", "Section")`. This is
telling us that these files are being joined on the basis of all four of
these variables matching in both data sets; in other words, for rows to
be joined, they must match identically on all four of these variables.

This is related to not specifying anything for the `by =` part of the
function; by default, `full_join()` (and `left_join()`) will consider
any character strings with identical names that are present in both data
sets to be keys. But, it's generally better practice to specify the
variables on which we are joining.

#### [**Your Turn**]{style="color: green;"} **⤵**

In the code below, write your join like above, but add the
`by = c("student_id", "Course", "Subject", "Section")` part to your
code. You may notice the red message you may have noticed does not
appear. This is generally a better practice because you know precisely
on which variables you data sets are joining.

```{r}
# join together the gradebook and log_wrangled
joined_data <- full_join()

joined_data
```

**Hint:** If you're curious about how to format the use of the `by =`
part of your code, look up above at how you used this argument to the
`full_join()` function.

What do you notice about the result---the data you joined? In
particular, how does it differ from the two data sets from which it was
created? Add one or more notes below.

-   

## 3. EXPLORE

Having prepared both of the data sets we joined together, and worked
hard to join those data sets, we're now ready to use this joined data
set in our exploration of how the time students spent on the course LMS
relate to the number of points they earned throughout the course.

We'll be using the {ggplot2} package again, but, this time, will be
creating a different type of plot.

Run the code below to create a scatter plot of the proportion of points
students earned and the number of hours they spent on the course LMS.

```{r}
ggplot(joined_data, aes(x = time_spent_hours, y = proportion_earned)) +
  geom_point()
```

What do you notice about this graph? And what do you wonder? How about
the code---what do you notice about it (and what do you wonder)? Add one
or more of what you see as the most important elements.

-   

Using {ggplot2} makes it efficient to iterate through different versions
of similar plots. For instance, we can color the points by a third
variable, such as the reason for which students enrolled in the course,
to begin to explore what was going on for students who spent very little
time on the course:

```{r}
ggplot(joined_data, aes(x = time_spent_hours, y = proportion_earned, color = enrollment_status)) +
  geom_point()
```

#### [**Your Turn**]{style="color: green;"} **⤵**

We can also additionally create faceted plots, like the one you created
in the last learning lab. In the code below, facet the plot by
`Subject`.

```{r}
ggplot(joined_data, aes(x = time_spent_hours, y = proportion_earned, color = enrollment_status)) +
  geom_point()
```

**Hint**: Take a peak at the code you wrote for the first learning lab
in which you created a faceted plot.

You may wish to style your plot. A few ways you can do that are as
follows; we'll discuss more throughout the institute. For each of the
following, add them to your plot by adding a plus symbol to the line
prior to the line you are adding. For instance, the following code
styles the x-axis label of a plot:

```{r}
ggplot(joined_data, aes(x = time_spent_hours, y = proportion_earned, color = enrollment_status)) +
  geom_point() +
  xlab("Time Spent (Hours)")
```

#### [**Your Turn**]{style="color: green;"} **⤵**

Try adding (and modifying, if you'd like) any of the following to the
faceted plot you created in the code chunk below:

-   `xlab("Time Spent (Hours)")`

-   `ylab("Proportion of Points Earned")`

-   `scale_color_brewer("Enrollment Status", type = "qual", palette = 3)`

-   `ggtitle("How Time Spent on Course LMS is Related to Points Earned in the Course")`

-   `theme(legend.position = "bottom")`

```{r}

```

Once you have settled on a plot you are happy with (for now!), add a
sentence or two interpreting your graph (like you were describing it
within a manuscript):

-   

## 5. COMMUNICATE

In this learning lab, we covered a lot of ground, from extensively
wrangling our data (further), joining together two data sets, and
creating plots that depicted the relations between two variables. We
also built on the ggplot'ting we began in the first learning lab, going
deeper in terms of the the type of plot we created as well as how we
styled it.

Like for the first learning lab, below, add a few notes in response to
the prompts:

*One thing I took away from this learning lab*:

*One thing I want to learn more about*:

## Reach

For this learning lab, your reach is to join together two of your own
data sets. Even if you don't expand on this code here, gaining
experience joining data sets with which you're familiar can be handy. If
you do this, re-knit your document when complete or at a stopping point
so you have a record of your work!

If you do not have data you wish to join at this time, consider reading
one of these resources on joining or that involve joining:

Relational Data from the excellent R for Data Science book:
<https://r4ds.had.co.nz/relational-data.html>

Walkthrough 1: The Education Data Science Pipeline With Online Science
Class Data from Data Science in Education Using
R<https://datascienceineducation.com/c07.html>
