---
title: "Learning Analytics Foundations Learning Lab 4"
output:
  html_notebook: default
  html_document:
    df_print: paged
    toc: true
  pdf_document: default
name: ''
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The culmination of this module involves using data from all three data
sets---on the time students spent on the course LMS, their academic
achievement, and their self reports---in a model, particularly, a linear
(regression) model to understand differences in students' academic
achievement. In addition to considering how to fit and interpret such a
model in R, we create and tailor a report that we can share with
collaborators and others. Let's begin!

## 1. PREPARE

### Packages 📦

#### [**Your Turn**]{style="color: green;"} **⤵**

Below, load the {tidyverse} package.

```{r}

```

### Data

We'll load the `data-to-explore` data we worked with in the prior
learning lab.

Below, read that file, saved in the `/data` directory, and assign it the
name `data_to_explore`.

```{r}

```

**Hint**: Refer to how we read data files in the last few learning labs
for a template or starting point for your code.

Before we proceed, let's take a step we so often take: taking a look at
our data. We haven't emphasized this function very much because it takes
you as the R programmer out of your environment---it opens a new window.
Nevertheless, it is good to have the `View()` function as a part of your
toolkit. To view the data frame you loaded, run `View(data_to_explore)`
in the console. Yes, the "V" is capitalized---very unusual for an R
function. Because this function is a bit atypical in more ways than one,
we have two recommendations concerning its use:

-   Use it strictly in the console. Because it opens a new viewing
    window, including it in an R Markdown script can cause issues when
    rendering an HTML (or PDF) report.

-   Close the viewer window that opens once you have viewed the data.
    Keeping it open can clutter your work space a bit and can lead to
    confusion about what data frame it was you viewed.

## 2. WRANGLE

We often think of wrangling as a step *prior to* exploring and modeling
data. However, we often carry out a cycle involving wrangling,
exploring, and modeling. For instance, we may explore our data to find
that a variable is not prepared or cleaned how it needs to be to achieve
our aim. Or, we may estimate a model, but find that the levels of the
qualitative (or categorical) variables are not encoded in such a way to
allow us to estimate the things we wish to (such as which levels differ
from a reference group).

If some of this modeling jargon is new, then you're still in the right
spot! Let's proceed to some concrete examples of how wrangling and
modeling, particularly, are related to one another.

## 4. MODEL

`data_to_explore` is a fairly large data frame, consisting of around
1,000 observations and around 15 columns. For the purpose of this
learning lab, let's consider the `proportion_earned` variable to be our
dependent, or the outcome, variable.

You may be new to linear regression models, or you may have a lot of
experience. In brief, a linear regression model involves estimating the
relationships between one or more *independent variables* with one
dependent variable. Mathematically, it can be written like the
following.

$$
\operatorname{dependentvar} = \beta_{0} + \beta_{1}(\operatorname{independentvar}) + \epsilon
$$

Here, the `dependentvar` is predicted by two *coefficients*, or things
that help to explain the dependent variable. The first coefficient,
$\beta_0$, is the intercept. This coefficient tells us what the
estimated value of the dependent variable is when the independent
variable (`independentvar`) is equal to 0. The other coefficient,
$\beta_1$, or the slope, represents the association of a one-unit change
in the independent variable in the value of the dependent variable.

Let's consider a concrete example. We'll use the `lm()` function in R to
estimate a linear regression model.

The following code estimates a model in which `proportion_earned`, the
proportion of points students earned, is the dependent variable. It is
predicted by one independent variable, `int`, students' self-reported
interest in science.

```{r}
lm(proportion_earned ~ int, data = data_to_explore)
```

Let's take a look at the output.

We can see that the intercept is estimated at 0.53. This tells us that
when students' interest is equal to zero, their predicted proportion of
points earned is 0.53---not such a great grade! But, for every one-unit
increase in students' interest in science, their estimate proportion of
points earned was 0.055. Let's consider what the mean value is for
students' interest.

The following is some new code, but we'll elaborate on the use of
`summarize()` later. For now, know that it's a way to create summary
statistics such as the mean, but also the standard deviation, or the
minimum or maximum of a value. At its core, think of `summarize()` as a
function that returns a single value (whether it's a mean, median,
standard deviation---whichever!) that summarizes a single column.

The argument `na.rm = TRUE` tells R that it can ignore missing, or `NA`
values, and to calculate the summary statistic using the non-missing
values.

```{r}
data_to_explore %>% 
  summarize(mean_interest = mean(int, na.rm = TRUE))
```

The mean value for interest is quite high. If we multiple the estimate
relationship between interest and proportion of points
earned---0.055---by this, the mean interest across all of the
students---we can determine that students' estimate final grade was
0.055 X 4.3, or 0.236, which we can add to the intercept, 0.535, which
equals 0.771. In other words, the student with average interest in
science earned 0.771 of possible points in the group---not associated
with a great grade, but not bad!

A linear regression model builds on this intuition by allowing us to add
variables other than students' interest to understand and explain the
points they earned in the course.

We can add additional predictor variables by separating variables with a
plus symbol.

```{r}
lm(proportion_earned ~ int + time_spent_hours, data = data_to_explore)
```

Let's add a third variable, this time a categorical variable, one for
students' gender as recorded in the LMS. We note that this is a course
variable, one inferior to one that we would have had had we asked
students to report their gender.

```{r}
lm(proportion_earned ~ int + time_spent_hours + gender, data = data_to_explore)
```

In the output, we can see that there is now a fourth coefficient,
`genderM`. This variable indicates the estimate difference between male
and female students. Thus, this effect of -0.009 can be considered to be
the "male effect" in comparison to females: Males earn an estimate
(nearly) a 0.01 proportion points fewer than females. Imagine we want to
consider, instead, the "female" effect.

The following code treats `gender` as a factor, but makes `M`, for
males, to be the reference group.

```{r}
data_to_explore <- data_to_explore %>% 
  mutate(gender = as.factor(gender),
         gender = fct_relevel(gender, "M"))
```

Then, when we estimate the model, we can see now that the estimate
effect is `genderF`, telling us the estimated proportion of points
female students earned relative to males. Since there were only two
groups for gender (again, because of the way the LMS only reported
students' genders as male or female, a serious limitation of the data
that we would want to take care to report), the sign of the effect
simply flipped, but if we had multiple groups, carefully choosing the
reference group can make a difference in the interpretability of the
model.

```{r}
lm(proportion_earned ~ int + time_spent_hours + gender, data = data_to_explore)
```

One last step. We can save the output of the function to an
object---let's say `m1`, standing for model 1. We can then use the
summary function to view a much more feature-rich summary of the
estimated model.

```{r}
m1 <- lm(proportion_earned ~ int + time_spent_hours + gender, data = data_to_explore)
summary(m1)
```

#### [**Your Turn**]{style="color: green;"} **⤵**

Below, estimate at least two different regression models, viewing and
interpreting the model results. You'll have to fill a lot in to estimate
these models! Consider any and all of the variables, and whether you
need to wrangle them into shape to meaningfully interpret them.

```{r}
m2 <- lm()
summary(m2)
```

Add a note or two interpreting the above model (`m2`) next:

-   

```{r}
m3 <- lm()
summary(m3)
```

Add another or note or two interpreting the above model (`m3`) next:

-   

## 5. COMMUNICATE

In this learning lab, we focused on modeling our data. We also discussed
how to interpret a linear regression model, and how to interpret the
output of one estimated within R. For some, this may be brand new, for
others, a refresher of a topic in which you have substantial expertise.
This is a seriously deep topic, with many future directions you may wish
to take. You may be interested in multi-level models; estimating them in
R is very similar to estimating a linear model; a brief tutorial in
which you may be interested is here:
<https://datascienceineducation.com/c13.html>

To complete this learning lab, "knit" the document to HTML. View the
file; may may notice there is now a table of contents! This is from the
`toc: true` argument added to the very top of this file. Consider making
other modifications to the front matter based on the options detailed
here: <https://bookdown.org/yihui/rmarkdown/html-document.html>

Good work! Below, add a few notes in response to the prompts:

*One thing I took away from this learning lab*:

*One thing I want to learn more about*:

## Reach

As a reach, consider whether you can estimate a model using your own
data. 

If interested, follow the link above to a resource on estimating a multi-level model, and read about and/or explore estimating one with your own data.
